\notechapter{N-20220715}{Conics and the Dedekind Construction of Addition}

\section{Conic sections}

The note starts with the geometry of conic sections.  A cone cut by a plane produces the classical conics:
\[
  \text{circle},\quad \text{ellipse},\quad \text{parabola},\quad \text{hyperbola}.
\]
An ellipse can be defined as the locus of points \(P\) for which the sum of distances to two foci is constant:
\[
  PF_1+PF_2=2a.
\]
With foci \(F_1=(-c,0)\), \(F_2=(c,0)\), this gives the standard equation
\[
  \frac{x^2}{a^2}+\frac{y^2}{b^2}=1,\qquad c^2=a^2-b^2.
\]
The eccentricity is
\[
  e=\frac ca,\qquad 0<e<1.
\]

A hyperbola is defined by a constant difference of distances to the two foci:
\[
  |PF_1-PF_2|=2a,
\]
with standard equations
\[
  \frac{x^2}{a^2}-\frac{y^2}{b^2}=1
\]
or
\[
  \frac{y^2}{a^2}-\frac{x^2}{b^2}=1,
\]
depending on the transverse axis.

A parabola is the locus of points equidistant from a focus and a directrix.  Its standard equations are
\[
  y^2=2px,\qquad y^2=-2px,\qquad x^2=2py,\qquad x^2=-2py.
\]
For \(y^2=2px\), the focus is \((p/2,0)\), the directrix is \(x=-p/2\), and the eccentricity is \(1\).

\section{Latus rectum and tangent geometry}

For an ellipse, a vertical line through a focus intersects the ellipse in endpoints of a latus rectum.  The length is
\[
  \frac{2b^2}{a}.
\]
The tangent at \(P=(x_0,y_0)\) to
\[
  \frac{x^2}{a^2}+\frac{y^2}{b^2}=1
\]
is
\[
  \frac{xx_0}{a^2}+\frac{yy_0}{b^2}=1.
\]
The note also records the focus-directrix relation.  For an ellipse with eccentricity \(e=c/a\), the directrices are
\[
  x=\pm\frac{a^2}{c}.
\]
The ratio
\[
  \frac{PF}{\operatorname{dist}(P,\text{directrix})}=e
\]
characterizes the conic.

\section{Rational and irrational density}

The page then shifts to real analysis and proves that rational and irrational numbers are dense in \(\mathbb R\).  For rationals, given \(x\in\mathbb R\) and \(\varepsilon>0\), choose \(q\in\mathbb N\) with
\[
  \frac1q<\varepsilon.
\]
Then choose an integer \(p\) such that
\[
  p\le qx<p+1.
\]
Thus
\[
  \left|x-\frac pq\right|<\frac1q<\varepsilon.
\]
For irrationals, take a rational approximation \(r\) to \(x-\sqrt2/N\), then
\[
  r+\frac{\sqrt2}{N}
\]
is irrational and can be made arbitrarily close to \(x\).

\section{Vector spaces}

The note includes the vector space axioms.  Let \(F\) be a field and \(V\) a set equipped with addition and scalar multiplication:
\[
  +:V\times V\to V,\qquad \cdot:F\times V\to V.
\]
The axioms are associativity and commutativity of addition, existence of \(0\), additive inverses, compatibility of scalar multiplication, distributivity, and \(1v=v\).  With the usual coordinate operations,
\[
  \mathbb R^n
\]
is an \(\mathbb R\)-linear space.

\section{Dedekind cuts and order}

The note returns to Dedekind cuts.  We denote by \(\mathbb R\) the set of Dedekind cuts.  For cuts \(X,Y\), define an order by
\[
  X<Y\quad\Longleftrightarrow\quad X\subset Y,\quad X\neq Y.
\]
This order is total because if \(X\not\subseteq Y\), choose \(x\in X\setminus Y\); the downward-closed property of cuts forces \(Y\subset X\).

\section{Addition of Dedekind cuts}

Define
\[
  X+Y=\{x+y:x\in X,\ y\in Y\}.
\]
The zero cut is
\[
  \overline0=\{x\in\mathbb Q:x<0\}.
\]
One must check that \(X+Y\) is again a Dedekind cut:
\begin{enumerate}[(i)]
  \item it is nonempty and not all of \(\mathbb Q\);
  \item it is downward closed;
  \item it has no greatest element.

\end{enumerate}
The page gives the standard proof.  For downward closure, if \(z<x+y\), then \(z=x+(z-x)\) and \(z-x<y\), so \(z-x\in Y\).  For no greatest element, choose larger elements \(x'>x\) in \(X\) and \(y'>y\) in \(Y\); then \(x'+y'>x+y\).

Associativity and commutativity follow directly:
\[
  (X+Y)+Z=X+(Y+Z),\qquad X+Y=Y+X.
\]
Order compatibility is subtler.  One needs the lemma: if \(X\) is a cut and \(n\in\mathbb N\), then there exist \(x\in X\) and \(x'\in X'\) such that
\[
  0<x'-x<\frac1n.
\]
This approximation lemma lets one prove
\[
  X<Y\quad\Rightarrow\quad X+Z<Y+Z.
\]

\section{Additive inverses}

For a cut \(X\), define
\[
  -X=\{y-x':y\in\overline0,\ x'\in X'\}.
\]
Then \(-X\) is the unique cut \(Z\) such that
\[
  X+Z=\overline0.
\]
This reproduces the usual negative of a real number.  In particular, if \(p/q\in\mathbb Q\), then
\[
  -X_{p/q}=X_{-p/q}.
\]


\section{Conics through focus and directrix}

A conic can be defined by the ratio
\[
  e=\frac{\text{distance to focus}}{\text{distance to directrix}}.
\]
If \(e<1\), the conic is an ellipse.  If \(e=1\), it is a parabola.  If \(e>1\), it is a hyperbola.  This unifies the three classical curves by one parameter: eccentricity.

\section{Dedekind cuts and addition}

The Dedekind construction treats a real number as a cut of rational numbers.  Addition is then defined by adding representatives from cuts:
\[
  A+B=\{a+b:a\in A,\ b\in B\}.
\]
The hard part is not writing the formula, but proving that it again satisfies the properties of a cut.  This illustrates how arithmetic operations must be reconstructed after the real numbers are built.

\section{A compact bridge}

This note is a striking mixture: classical conics and the construction of real addition by Dedekind cuts.  The hidden connection is that both topics turn geometry into algebra.  A conic becomes an equation and a focus-directrix ratio; a real number becomes a cut with algebraic operations.


\paragraph{Expanded synthesis.}
For this chapter, the elementary material should be read as a study of what remains invariant when coordinates change. The earlier sections (`Conic sections`, `Latus rectum and tangent geometry`, `Rational and irrational density`, `Vector spaces`, `Dedekind cuts and order`) compute with arrays, bases, pivots, determinants, or eigenvectors; the deeper object is a linear map together with the spaces it acts on. A matrix is a representation of that map after bases have been chosen. This is why formulas such as
\[
  A' = P^{-1}AP,\qquad \widetilde A = T^{-1}AS
\]
are not cosmetic: they distinguish an intrinsic morphism from a coordinate description.

The structural hierarchy is as follows. Kernels record what a map forgets, images record what it reaches, cokernels record obstructions to solvability, and quotients record information after an equivalence has been imposed. Determinants act on the top exterior power and therefore measure oriented volume; traces are invariant under cyclic rearrangement and become characters in representation theory; adjoints depend on an inner product and lead to self-adjoint spectral theory.

A useful way to return to the computations is to ask, for every formula, which invariant it is measuring. Row reduction measures rank and kernel; diagonalization measures decomposition into invariant subspaces; Gram--Schmidt measures orthogonal projection; positive definiteness measures ellipsoid geometry; tensor notation measures how components transform. The elementary methods are therefore the finite-dimensional laboratory in which duality, quotienting, functoriality, and spectral decomposition can be seen clearly.


\section{Conics and cuts as two kinds of construction}

The conic part and the Dedekind-cut part look unrelated, but both are about turning an intuitive geometric object into an exact algebraic definition.  An ellipse is not just a picture; it is the set of points satisfying a focus-distance equation.  A real number is not just a point on a drawn line; it can be encoded as a cut in \(\mathbb Q\).

This is a recurring mathematical move: replace a mental image by a definition that can be manipulated.  Once that is done, one can prove tangent formulas for conics and algebraic laws for real numbers.
