\notechapter{N-20221126}{Congruence Classes, Complete Residue Systems, and Euler's Function}

\section{Congruence}

Let \(m\) be a positive integer and \(a,b\in\mathbb Z\).  We say that \(a\) and \(b\) are congruent modulo \(m\), written
\[
  a\equiv b\pmod m,
\]
if
\[
  m\mid(a-b).
\]
This definition is just divisibility language with a new notation.  It allows us to work with remainders rather than whole integers.

\section{Basic properties}

The note lists the standard properties:
\begin{enumerate}[(i)]
  \item Reflexivity: \(a\equiv a\pmod m\).
  \item Symmetry: if \(a\equiv b\pmod m\), then \(b\equiv a\pmod m\).
  \item Transitivity: if \(a\equiv b\pmod m\) and \(b\equiv c\pmod m\), then \(a\equiv c\pmod m\).
  \item Compatibility with addition and multiplication:
  \[
    a\equiv b,\ c\equiv d\pmod m
    \quad\Rightarrow\quad
    a+c\equiv b+d,\quad ac\equiv bd\pmod m.
  \]
  \item If \(a\equiv b\pmod m\), then
  \[
    a^n\equiv b^n\pmod m
  \]
  for every positive integer \(n\).
\end{enumerate}

The last two rules are what make congruence useful: one can replace a large number by a small remainder and keep computing.

\section{Residue classes}

A residue class modulo \(n\) is a set
\[
  M_i=\{x\in\mathbb Z:x\equiv i\pmod n\}.
\]
The integers split into \(n\) disjoint classes:
\[
  M_0,M_1,\ldots,M_{n-1}.
\]
This is a partition of \(\mathbb Z\).  The note records two important consequences:
\begin{enumerate}[(i)]
  \item among any \(n+1\) integers, two are congruent modulo \(n\);
  \item if \((i,n)=1\), then every integer in \(M_i\) is coprime to \(n\).

\end{enumerate}
The first is the pigeonhole principle; the second follows because congruent numbers have the same gcd with \(n\).

\section{Complete and reduced residue systems}

A complete residue system modulo \(n\) is any set of \(n\) representatives, one from each residue class.  The standard example is
\[
  0,1,2,\ldots,n-1.
\]
A reduced residue system modulo \(n\) consists of one representative from each residue class coprime to \(n\).  The number of such classes is Euler's function
\[
  \varphi(n).
\]
If
\[
  n=p_1^{\alpha_1}\cdots p_k^{\alpha_k},
\]
then
\[
  \varphi(n)=n\left(1-\frac1{p_1}\right)\cdots\left(1-\frac1{p_k}\right).
\]
In particular,
\[
  \varphi(p)=p-1,\qquad \varphi(p^a)=p^a-p^{a-1}.
\]

\section{Example: last digit}

To compute the last digit of \(15^{200}\), work modulo \(10\).  Since
\[
  15\equiv5\pmod{10},
\]
we get
\[
  15^{200}\equiv5^{200}\equiv5\pmod{10}.
\]
Thus the last digit is \(5\).

The note's method is to reduce early.  Do not expand powers; replace the base by its residue.

\section[Example: modulo 17]{Example: modulo \(17\)}

To compute
\[
  15^{100}\pmod{17},
\]
notice
\[
  15\equiv-2\pmod{17}.
\]
Then
\[
  15^{100}\equiv (-2)^{100}=2^{100}\pmod{17}.
\]
Since
\[
  2^4=16\equiv-1\pmod{17},
\]
we have
\[
  2^8\equiv1\pmod{17}.
\]
Therefore
\[
  2^{100}=2^{96}2^4\equiv1^{12}(-1)\equiv16\pmod{17}.
\]

\section{Example: reducing a long expression}

For an expression such as
\[
  2001\cdot2002+2003\cdot2004+2005\cdot2006
\]
modulo \(45\), reduce every factor modulo \(45\) first.  The arithmetic becomes small.  The main habit is:
\[
  \text{large expression}\quad\longrightarrow\quad\text{small residues}.
\]
This is the same principle behind all modular computation.

\section{A polynomial congruence by induction}

The note also includes a divisibility proof for an expression depending on \(k\), of the form
\[
  12^{2k+1}+36^k+5^k+3.
\]
The standard method is induction or modular periodicity.  One chooses the modulus, checks the base case, and then shows the expression changes by a multiple of the modulus when \(k\) increases.  Equivalently, reduce each base modulo the modulus and use periodicity of powers.

\section[Quadratic residues modulo 8]{Quadratic residues modulo \(8\)}

A consolidation exercise proves that a perfect square modulo \(8\) is congruent to
\[
  0,\ 1,\quad\text{or}\quad4.
\]
Indeed, every integer is congruent modulo \(8\) to one of
\[
  0,1,2,3,4,5,6,7.
\]
Squaring gives residues
\[
  0,1,4,1,0,1,4,1.
\]
Thus only \(0,1,4\) occur.  This is a powerful small filter in number theory problems.

\section{Where the idea leads}

Congruence classes are quotient thinking.  Instead of remembering the whole integer, we remember only its class in
\[
  \mathbb Z/n\mathbb Z.
\]
Complete residue systems are representatives of all classes; reduced residue systems are the invertible classes.  Euler's function counts the units.  This note is therefore the elementary entrance to the ring \(\mathbb Z/n\mathbb Z\) and its unit group.



\paragraph{Expanded synthesis.}
The number-theoretic content of this chapter is best understood as arithmetic seen through algebraic structure. The earlier sections (`Congruence`, `Basic properties`, `Residue classes`, `Complete and reduced residue systems`, `Example: last digit`) compute divisibility, congruences, residues, prime factorizations, or finite-field examples. The deeper language is ideals, quotient rings, unit groups, valuations, and field extensions.

Divisibility in $\mathbb Z$ is containment of principal ideals; congruence modulo $n$ is equality inside the quotient ring $\mathbb Z/n\mathbb Z$; invertibility modulo $n$ means being a unit; the Chinese remainder theorem is a product decomposition of quotient rings. Prime factorization supplies coordinates through valuations:
\[
  n=\prod_p p^{v_p(n)}.
\]
Gcd and lcm are coordinatewise minimum and maximum in these valuation coordinates.

This algebraic reading explains why elementary tricks scale upward. Euclidean algorithms become the theory of Euclidean domains and PIDs; divisor sums become Dirichlet convolution and Euler products; roots of unity become cyclotomic fields and Galois actions; finite polynomial arithmetic becomes finite-field theory. The original computations should therefore be kept as the algorithmic core, while the advanced language identifies the structure that makes the algorithms work.


\section{Residue systems as groups of units}

A reduced residue system modulo $n$ is the unit group
\[
  (\mathbb Z/n\mathbb Z)^\times.
\]
Its order is Euler's function $\varphi(n)$.  Euler's theorem says that if $\gcd(a,n)=1$, then
\[
  a^{\varphi(n)}\equiv1\pmod n.
\]
This is simply Lagrange's theorem applied to the finite group of units.

Thus the elementary reduced-residue computations are group theory in disguise.

\section{CRT decomposition of unit groups}

If $m,n$ are coprime, then
\[
  \mathbb Z/mn\mathbb Z\cong \mathbb Z/m\mathbb Z\times \mathbb Z/n\mathbb Z.
\]
Taking units gives
\[
  (\mathbb Z/mn\mathbb Z)^\times\cong
  (\mathbb Z/m\mathbb Z)^\times\times(\mathbb Z/n\mathbb Z)^\times.
\]
This explains the multiplicativity of $\varphi$.

For prime powers, the structure of the unit group is more subtle.  For odd primes $p$, the group $(\mathbb Z/p^k\mathbb Z)^\times$ is cyclic of order $p^{k-1}(p-1)$.  Powers of $2$ require separate treatment.

\section{Primitive roots and characters}

A primitive root modulo $n$ is a generator of the unit group.  When the unit group is cyclic, every unit is a power of one element.  This converts multiplicative congruence problems into exponent arithmetic.

Dirichlet characters are homomorphisms
\[
  \chi:(\mathbb Z/n\mathbb Z)^\times\to\mathbb C^\times
\]
extended to all integers by setting $\chi(a)=0$ when $\gcd(a,n)>1$.  They are Fourier characters on the finite abelian group of units and are central in Dirichlet's theorem on primes in arithmetic progressions.

\section{Returning to residue-system exercises}

The note's residue tables, reduced systems, and Euler-function formulas are not just modular arithmetic.  They are computations inside finite groups.  The next levels are group decomposition, primitive roots, and characters.

% Second-pass deepening for waves 2-4: wave 3 A-C part 3
\section{Primitive roots as logarithms in finite groups}

When $(\mathbb Z/n\mathbb Z)^\times$ is cyclic with generator $g$, every unit has the form $g^k$.  This turns multiplication into addition of exponents modulo the group order.  In other words, a primitive root supplies a discrete logarithm coordinate.

This is powerful but also limited: not every modulus has primitive roots.  The classification says primitive roots exist for
\[
  n=1,2,4,p^k,2p^k
\]
where $p$ is an odd prime.  This explains why some modular arithmetic problems become simple after choosing a primitive root, while others require decomposing the unit group into several cyclic factors.

\section{Characters as Fourier analysis on unit groups}

Dirichlet characters are not arbitrary functions; they are group homomorphisms from the unit group to complex roots of unity.  Orthogonality of characters gives filters for congruence classes.  This is the multiplicative analogue of using roots of unity to filter residues modulo $n$.

The elementary reduced-residue system is therefore the domain on which finite harmonic analysis is performed.
