\notechapter{N-20251210}{Calculus Review II: Monotonicity, Integration Methods, Applications, and More}



\section{Strict monotonicity from the derivative}

If $f$ is differentiable on an interval $I$ and
\[
  f'(x)>0\quad(x\in I),
\]
then $f$ is strictly increasing on $I$.  If
\[
  f'(x)<0\quad(x\in I),
\]
then $f$ is strictly decreasing.

The proof uses the Lagrange mean value theorem.  For $x_2>x_1$,
\[
  f(x_2)-f(x_1)=f'(\xi)(x_2-x_1)
\]
for some $\xi\in(x_1,x_2)$.  If $f'(\xi)>0$, then the difference is positive.

\section{First derivative test for extrema}

Suppose $f$ is continuous near $x_0$ and differentiable in the punctured neighborhood.  If the sign of $f'$ changes from negative to positive at $x_0$, then $f(x_0)$ is a local minimum.  If the sign changes from positive to negative, then $f(x_0)$ is a local maximum.

The page records the compact criterion:
\[
  (x-x_0)f'(x)>0
  \quad\Longrightarrow\quad
  f(x_0)	ext{ is a local minimum},
\]
\[
  (x-x_0)f'(x)<0
  \quad\Longrightarrow\quad
  f(x_0)	ext{ is a local maximum}.
\]

\section{Second derivative test}

If
\[
  f'(x_0)=0,\qquad f''(x_0)\ne0,
\]
then
\[
  f''(x_0)>0\quad\Longrightarrow\quad x_0	ext{ is a local minimum},
\]
\[
  f''(x_0)<0\quad\Longrightarrow\quad x_0	ext{ is a local maximum}.
\]
The handwritten note adds the mnemonic that $f''>0$ looks like a smile and $f''<0$ looks like a frown.

\section{Concavity and inflection points}

A curve segment is concave up if it lies above its tangent lines, and concave down if it lies below its tangent lines.  In the usual calculus language,
\[
  f''(x)>0
\]
indicates concave up, while
\[
  f''(x)<0
\]
indicates concave down.

An inflection point is a point where the concavity changes.  A necessary condition, when $f''$ exists, is
\[
  f''(x_0)=0.
\]
But this is not sufficient.  The note highlights that $f''(x_0)=0$ or nonexistence of $f''(x_0)$ only gives candidates; one must check whether the sign of $f''$ changes across $x_0$.

The scanned page distinguishes several types: smooth horizontal inflection, smooth non-horizontal inflection, and sharp or vertical tangent behavior.  The visual test is concavity change, not merely whether the tangent exists.

\section{Asymptotes}

The notes classify asymptotes.

A vertical asymptote occurs when
\[
  \lim_{x\to x_0}f(x)=\pm\infty.
\]
Then $x=x_0$ is a vertical asymptote.

A horizontal asymptote occurs when
\[
  \lim_{x\to+\infty}f(x)=c
  \quad	ext{or}\quad
  \lim_{x\to-\infty}f(x)=c.
\]
Then $y=c$ is a horizontal asymptote.

For an oblique asymptote
\[
  y=ax+b,
\]
we require
\[
  \lim_{x\to\infty}[f(x)-(ax+b)]=0.
\]
The page derives
\[
  a=\lim_{x\to\infty}
\frac{f(x)}x,
\]
and then
\[
  b=\lim_{x\to\infty}(f(x)-ax).
\]
These formulas are highlighted in the source.

\section{General procedure for sketching a curve}

The pages give a curve-sketching checklist:

\begin{enumerate}[(i)]
  \item determine the domain;
  \item check symmetry and periodicity;
  \item compute $f'(x)$ and $f''(x)$;
  \item find critical points and possible inflection points;
  \item determine intervals of monotonicity and concavity;
  \item find asymptotes;
  \item locate intercepts or special points if useful;
  \item draw the graph according to this information.
\end{enumerate}

This is not a mechanical ritual.  The derivative table tells how the curve moves; the second derivative table tells how it bends; asymptotes determine the large-scale shape.

\section{Basic antiderivative table}

The integration section records standard formulas:
\[
  \int 0\,dx=C,\qquad
  \int x^\mu\,dx=
\frac{x^{\mu+1}}{\mu+1}+C\quad(\mu\ne-1),
\]
\[
  \int
\frac1x\,dx=\ln|x|+C,
  \qquad
  \int a^x\,dx=
\frac{a^x}{\ln a}+C.
\]
Trigonometric formulas include
\[
  \int\sin x\,dx=-\cos x+C,\qquad
  \int\cos x\,dx=\sin x+C,
\]
\[
  \int\sec^2x\,dx=	an x+C,\qquad
  \int\csc^2x\,dx=-\cot x+C,
\]
\[
  \int
\frac{dx}{\sqrt{1-x^2}}=\arcsin x+C,
  \qquad
  \int
\frac{dx}{1+x^2}=\arctan x+C.
\]
The page also lists logarithmic forms such as
\[
  \int
\frac{dx}{\sqrt{x^2+a^2}}
  =\ln|x+\sqrt{x^2+a^2}|+C.
\]

\section{Substitution}

For a composite function, the substitution rule is
\[
  \int f(
arphi(v))
arphi'(v)\,dv
  =
  \int f(u)\,du,
  \qquad u=
arphi(v).
\]
Equivalently,
\[
  \int f(u)\,du
  =
  \int f(
arphi(v))
arphi'(v)\,dv.
\]
The scanned page calls the two directions the first and second substitution methods.  The first recognizes a derivative already present; the second introduces a new variable to simplify the integral.

\section{Integration by parts}

From
\[
  d(uv)=u\,dv+v\,du,
\]
one obtains
\[
  \int u\,dv=uv-\int v\,du.
\]
The page presents integration by parts as the other major method after substitution.  Its use depends on choosing $u$ so that differentiating it simplifies the expression and choosing $dv$ so that it can be integrated.

\section{Rational functions and partial fractions}

A rational function is
\[
\frac{P(x)}{Q(x)},
\]
where $P,Q$ are polynomials and $Q\ne0$.  If $\deg P\ge \deg Q$, first divide polynomials to write it as a polynomial plus a proper rational function.

The pages list the elementary partial fraction forms.  A real linear factor $(x-a)^k$ contributes
\[
\frac{A_1}{x-a}+
\frac{A_2}{(x-a)^2}+\cdots+
\frac{A_k}{(x-a)^k}.
\]
An irreducible quadratic factor $(x^2+px+q)^k$ with $p^2-4q<0$ contributes
\[
\frac{M_1x+N_1}{x^2+px+q}
  +\cdots+
\frac{M_kx+N_k}{(x^2+px+q)^k}.
\]
Thus integration of rational functions reduces to two basic integral types:
\[
  \int
\frac{dx}{(x-a)^k},
  \qquad
  \int
\frac{Mx+N}{(x^2+px+q)^k}\,dx.
\]

\section{Quadratic denominator reduction}

For the second type, split the numerator as a derivative part plus a remainder:
\[
  Mx+N=
\frac M2(2x+p)+\left(N-
\frac M2p
\right).
\]
Then
\[
  \int
\frac{Mx+N}{(x^2+px+q)^k}\,dx
  =
\frac M2 I_k^1+\left(N-
\frac M2p
\right)I_k^2,
\]
where
\[
  I_k^1=\int
\frac{d(x^2+px+q)}{(x^2+px+q)^k},
\]
and
\[
  I_k^2=\int
\frac{dx}{(x^2+px+q)^k}.
\]
Complete the square by setting
\[
  t=x+
\frac p2,\qquad
  a^2=q-
\frac{p^2}{4}>0.
\]
Then
\[
  x^2+px+q=t^2+a^2.
\]
For $k=1$,
\[
  I_1^2=
\frac1a\arctan
\frac ta+C.
\]
For higher $k$, the recurrence shown on the page is
\[
  I_{k+1}^2=
\frac1{2ka^2}
  \left(
\frac{t}{(t^2+a^2)^k}+(2k-1)I_k^2
\right).
\]

\section{Trigonometric rational functions}

The notes next handle rational functions of $\sin x$ and $\cos x$.  Use the tangent half-angle substitution
\[
  t=	an
\frac x2,\qquad -\pi<x<\pi.
\]
Then
\[
  \sin x=
\frac{2t}{1+t^2},\qquad
  \cos x=
\frac{1-t^2}{1+t^2},\qquad
  	an x=
\frac{2t}{1-t^2},
\]
and
\[
  dx=
\frac{2\,dt}{1+t^2}.
\]
Thus
\[
  \int R(\sin x,\cos x)\,dx
  =
  \int R\left(
\frac{2t}{1+t^2},
\frac{1-t^2}{1+t^2}
\right)
\frac{2\,dt}{1+t^2},
\]
which becomes a rational integral in $t$.

\section{Definite integrals and Riemann sums}

The page defines the definite integral.  Divide $[a,b]$ by
\[
  a=x_0<x_1<\cdots<x_n=b,
\]
let
\[
  \Delta x_i=x_i-x_{i-1},
\]
and choose sample points $\xi_i\in[x_{i-1},x_i]$.  The Riemann sum is
\[
  \sum_{i=1}^n f(\xi_i)\Delta x_i.
\]
If the limit exists as the mesh
\[
  \lambda=\max_i\Delta x_i
\]
tends to zero, then
\[
  \int_a^b f(x)\,dx
\]
is the definite integral.

The pages list sufficient conditions for integrability: continuous functions, monotone bounded functions, and bounded functions with finitely many discontinuities are integrable on closed intervals.

\section{Basic properties of definite integrals}

The standard properties are:
\[
  \int_a^b [\alpha f(x)+\beta g(x)]\,dx
  =
  \alpha\int_a^b f(x)\,dx+\beta\int_a^b g(x)\,dx,
\]
\[
  \int_a^b f(x)\,dx=-\int_b^a f(x)\,dx,\qquad
  \int_a^a f(x)\,dx=0,
\]
\[
  \int_a^b f(x)\,dx=\int_a^c f(x)\,dx+\int_c^b f(x)\,dx.
\]
If $f(x)\ge0$ on $[a,b]$, then
\[
  \int_a^b f(x)\,dx\ge0.
\]
If $f\le g$, then
\[
  \int_a^b f(x)\,dx\le\int_a^b g(x)\,dx.
\]

\section{Integral mean value theorem}

If $f$ is continuous on $[a,b]$, then there exists $\xi\in[a,b]$ such that
\[
  \int_a^b f(x)\,dx=f(\xi)(b-a).
\]
If $g$ does not change sign, the weighted form is
\[
  \int_a^b f(x)g(x)\,dx
  =
  f(\xi)\int_a^b g(x)\,dx.
\]
The note uses this to interpret integrals as average values.

\section{Applications: area, volume, and arc length}

Area between curves is computed by
\[
  A=\int_a^b [f(x)-g(x)]\,dx
\]
when $f\ge g$.  If horizontal slicing is better,
\[
  A=\int_c^d [x_{right}(y)-x_{left}(y)]\,dy.
\]

The volume of a solid of revolution around the $x$-axis is
\[
  V=\pi\int_a^b [f(x)]^2\,dx.
\]
The shell method has the form
\[
  V=2\pi\int_a^b x f(x)\,dx
\]
when rotating a region around the $y$-axis.

Arc length of $y=f(x)$ is
\[
  L=\int_a^b \sqrt{1+[f'(x)]^2}\,dx.
\]
For a parametric curve $x=x(t),y=y(t)$,
\[
  L=\int_\alpha^\beta
  \sqrt{(x'(t))^2+(y'(t))^2}\,dt.
\]

\section{Improper integrals}

The last pages include reminders about improper integrals.  If the interval is unbounded,
\[
  \int_a^{+\infty}f(x)\,dx
  =
  \lim_{R\to+\infty}\int_a^R f(x)\,dx.
\]
If the integrand has a singularity at $b$, then
\[
  \int_a^b f(x)\,dx
  =
  \lim_{t\to b^-}\int_a^t f(x)\,dx.
\]
Convergence depends on the existence of these limits.

A standard comparison is the $p$-integral:
\[
  \int_1^\infty 
\frac{dx}{x^p}
\]
converges iff $p>1$, while
\[
  \int_0^1 
\frac{dx}{x^p}
\]
converges iff $p<1$.

\section{Broader perspective}

This note connects derivative information with the global shape of a curve, then connects antiderivatives with accumulation.  Derivatives decide monotonicity, extrema, concavity, inflection points, and asymptotes.  Integration methods reduce complicated antiderivatives to known forms.  Definite integrals turn limits of sums into area, volume, length, and average value.  The same calculus ideas are being used both for qualitative geometry and quantitative computation.


\paragraph{Expanded synthesis.}
For this chapter, the elementary material should be read as a study of what remains invariant when coordinates change. The earlier sections (`Strict monotonicity from the derivative`, `First derivative test for extrema`, `Second derivative test`, `Concavity and inflection points`) compute with arrays, bases, pivots, determinants, or eigenvectors; the deeper object is a linear map together with the spaces it acts on. A matrix is a representation of that map after bases have been chosen. This is why formulas such as
\[
  A' = P^{-1}AP,\qquad \widetilde A = T^{-1}AS
\]
are not cosmetic: they distinguish an intrinsic morphism from a coordinate description.

The structural hierarchy is as follows. Kernels record what a map forgets, images record what it reaches, cokernels record obstructions to solvability, and quotients record information after an equivalence has been imposed. Determinants act on the top exterior power and therefore measure oriented volume; traces are invariant under cyclic rearrangement and become characters in representation theory; adjoints depend on an inner product and lead to self-adjoint spectral theory.

A useful way to return to the computations is to ask, for every formula, which invariant it is measuring. Row reduction measures rank and kernel; diagonalization measures decomposition into invariant subspaces; Gram--Schmidt measures orthogonal projection; positive definiteness measures ellipsoid geometry; tensor notation measures how components transform. The elementary methods are therefore the finite-dimensional laboratory in which duality, quotienting, functoriality, and spectral decomposition can be seen clearly.
