\notechapter{N-20230123}{Definite Integrals, Line Integrals, Work, Green-Type Thinking, and Vector Operations}

\section{Indefinite and definite integrals}

The note begins by distinguishing indefinite and definite integrals.  An indefinite integral is an antiderivative family:
\[
  \int f(x)\,dx=F(x)+C,\qquad F'(x)=f(x).
\]
A definite integral is a number:
\[
  \int_a^b f(x)\,dx.
\]
Geometrically, it is signed area under the graph.  Analytically, it is the limit of Riemann sums.

The fundamental theorem of calculus connects the two:
\[
  \int_a^b f(x)\,dx=F(b)-F(a)
\]
when $F'=f$.

\section{Area by slicing}

The page includes diagrams of shaded regions.  The standard method is to slice a region into thin strips.  If vertical strips are convenient, then
\[
  A=\int_a^b (y_{top}(x)-y_{bottom}(x))\,dx.
\]
If horizontal strips are more convenient, then
\[
  A=\int_c^d (x_{right}(y)-x_{left}(y))\,dy.
\]
The note's sketches show that choosing the direction of slicing is part of the solution, not a fixed rule.

\section{Work as an integral}

The physical meaning of a definite integral appears in work.  If a force $F(x)$ acts along a line, then the work from $x=a$ to $x=b$ is
\[
  W=\int_a^b F(x)\,dx.
\]
If the force is constant, this reduces to $W=F(b-a)$.  If the force changes with position, integration adds the small contributions
\[
  dW=F(x)\,dx.
\]
This is the same accumulation idea as area, but with a physical interpretation.

\section{Line integrals}

The note then introduces line integrals.  If a curve is parametrized by
\[
  r(t)=(x(t),y(t)),\qquad a\le t\le b,
\]
then the scalar line element is
\[
  ds=\sqrt{(x'(t))^2+(y'(t))^2}\,dt.
\]
A scalar line integral is
\[
  \int_C f\,ds
  =\int_a^b f(x(t),y(t))\sqrt{(x'(t))^2+(y'(t))^2}\,dt.
\]
For a vector field $F=(P,Q)$, the work integral is
\[
  \int_C F\cdot dr
  =\int_C P\,dx+Q\,dy
  =\int_a^b [P(x(t),y(t))x'(t)+Q(x(t),y(t))y'(t)]\,dt.
\]

\section{Green-type intuition}

The handwritten diagrams show a small oriented boundary and a vector field.  The idea is that circulation around a boundary can be related to a double integral over the region.  In the plane, Green's theorem says
\[
  \oint_{\partial D} P\,dx+Q\,dy
  =
  \iint_D \left(\frac{\partial Q}{\partial x}-\frac{\partial P}{\partial y}\right)\,dA.
\]
Even if the note does not prove the theorem fully, it records the key intuition: local rotation accumulated over an area produces circulation on the boundary.

\section{Polar coordinates}

The note uses polar coordinates in examples.  The change of variables is
\[
  x=r\cos\theta,\qquad y=r\sin\theta.
\]
The area element becomes
\[
  dA=r\,dr\,d\theta.
\]
The factor $r$ is the Jacobian.  Geometrically, a small polar rectangle has side lengths approximately $dr$ and $r\,d\theta$.

Thus a disk of radius $R$ has area
\[
  \int_0^{2\pi}\int_0^R r\,dr\,d\theta=\pi R^2.
\]

\section{Spherical coordinates}

The later page contains a spherical-coordinate diagram.  A common convention is
\[
  x=\rho\sin\varphi\cos\theta,\qquad
  y=\rho\sin\varphi\sin\theta,\qquad
  z=\rho\cos\varphi.
\]
The volume element is
\[
  dV=\rho^2\sin\varphi\,d\rho\,d\varphi\,d\theta.
\]
The factor $\rho^2\sin\varphi$ comes from the three small lengths:
\[
  d\rho,\qquad \rho\,d\varphi,\qquad \rho\sin\varphi\,d\theta.
\]
This explains the diagram on the final page: the small spherical patch has dimensions determined by radius and angle.

\section{Dot product and cross product}

The note also summarizes vector operations.  The dot product is
\[
  a\cdot b=|a||b|\cos\theta.
\]
It measures projection and appears in work:
\[
  dW=F\cdot dr.
\]
The cross product has magnitude
\[
  |a\times b|=|a||b|\sin\theta
\]
and direction given by orientation.  It measures the area of the parallelogram spanned by $a$ and $b$.

In coordinates,
\[
  a\times b=
  \begin{vmatrix}
  i&j&k\\
  a_1&a_2&a_3\\
  b_1&b_2&b_3
  \end{vmatrix}.
\]

\section{Expanded synthesis and advanced context}

The note develops integration as accumulation in increasingly geometric forms.  Ordinary definite integrals accumulate along a line.  Line integrals accumulate along a curve.  Double and triple integrals accumulate over area and volume.  Coordinate changes require Jacobians.  Dot products, cross products, and differential elements provide the language for physical quantities such as work, flux, and circulation.



\paragraph{Expanded synthesis.}
For this chapter, the elementary material should be read as a study of what remains invariant when coordinates change. The earlier sections (`Indefinite and definite integrals`, `Area by slicing`, `Work as an integral`, `Line integrals`, `Green-type intuition`) compute with arrays, bases, pivots, determinants, or eigenvectors; the deeper object is a linear map together with the spaces it acts on. A matrix is a representation of that map after bases have been chosen. This is why formulas such as
\[
  A' = P^{-1}AP,\qquad \widetilde A = T^{-1}AS
\]
are not cosmetic: they distinguish an intrinsic morphism from a coordinate description.

The structural hierarchy is as follows. Kernels record what a map forgets, images record what it reaches, cokernels record obstructions to solvability, and quotients record information after an equivalence has been imposed. Determinants act on the top exterior power and therefore measure oriented volume; traces are invariant under cyclic rearrangement and become characters in representation theory; adjoints depend on an inner product and lead to self-adjoint spectral theory.

A useful way to return to the computations is to ask, for every formula, which invariant it is measuring. Row reduction measures rank and kernel; diagonalization measures decomposition into invariant subspaces; Gram--Schmidt measures orthogonal projection; positive definiteness measures ellipsoid geometry; tensor notation measures how components transform. The elementary methods are therefore the finite-dimensional laboratory in which duality, quotienting, functoriality, and spectral decomposition can be seen clearly.


\section{Elementary-to-advanced bridge: line integrals as pullbacks}

For a parametrized curve $\gamma:[a,b]\to\mathbb R^n$, a line integral of a one-form
\[
  \omega=P_1dx_1+\cdots+P_ndx_n
\]
is
\[
  \int_\gamma \omega
  =\int_a^b \sum_i P_i(\gamma(t))\gamma_i'(t)\,dt.
\]
This is the pullback of the one-form to the interval:
\[
  \gamma^*\omega=\sum_i P_i(\gamma(t))\gamma_i'(t)dt.
\]
Thus the elementary formula $Pdx+Qdy$ is already differential-form notation.

\section{Elementary-to-advanced bridge: exact forms and potentials}

A one-form $\omega$ is exact if $\omega=df$ for some potential $f$.  Then
\[
  \int_\gamma df=f(\gamma(b))-f(\gamma(a)).
\]
This is the fundamental theorem of line integrals.

A form is closed if $d\omega=0$.  In the plane, for $\omega=Pdx+Qdy$, this means
\[
  Q_x-P_y=0.
\]
On simply connected domains, closed one-forms are exact.  On domains with holes, this may fail.

\section{Elementary-to-advanced bridge: conservation laws}

In mechanics, work is a line integral
\[
  W=\int_C F\cdot dr.
\]
If $F=-\nabla V$ is conservative, then
\[
  W=V(A)-V(B)
\]
depends only on endpoints.  Conservation of energy is therefore a statement about exact one-forms.

This connects elementary work integrals to topology: path independence is not only about derivatives; it also depends on whether the domain has holes.

\section{Elementary-to-advanced bridge: de Rham preview}

De Rham cohomology measures the failure of closed forms to be exact:
\[
  H^1_{dR}(M)=\frac{\{\text{closed 1-forms}\}}{\{\text{exact 1-forms}\}}.
\]
For the punctured plane, the form
\[
  \omega=\frac{-y\,dx+x\,dy}{x^2+y^2}
\]
is closed but not exact.  Its integral around the unit circle is $2\pi$.

The elementary line integral around a circle is therefore a measurement of a topological hole.

% Second-pass deepening for waves 2-4: wave 4 part 2
\section{Second-pass deepening: work integrals and gauge freedom}

If a force field is conservative, $F=-\nabla V$, then work depends only on endpoints.  The potential $V$ is not unique: adding a constant changes nothing.  This is the simplest form of gauge freedom.  The physically meaningful object is the differential $dV$, not the absolute zero of potential.

In electromagnetism, potentials have richer gauge freedom.  A vector potential $A$ may be changed by a gradient without changing the magnetic field $B=\nabla\times A$.  The elementary fact that potentials are determined only up to constants is the first example of this broader principle.

\section{Second-pass deepening: polar and spherical coordinates as adapted charts}

The original integral examples using polar or spherical coordinates should be read as choosing charts adapted to symmetry.  A chart is a coordinate map from a parameter domain to the geometric region.  The integration element is obtained by pulling back the metric or volume form.

Thus polar coordinates are not merely a trick for circular regions.  They are a coordinate chart on the punctured plane adapted to the action of rotations.  Spherical coordinates are adapted to radial symmetry in three dimensions.
