\notechapter{N-20220412}{Linear Algebra Continued: Indices, Rank, Transpose, Hermitian Conjugate, and Trace}

\section{Unit vectors and Kronecker delta}

This note follows the previous linear algebra material and collects notation that will be useful later.

\begin{definition}
A unit vector is a vector of length \(1\).  I write such a vector as \(\mathbf v\) when the direction matters more than its magnitude.
\end{definition}

\begin{definition}[Kronecker delta]
The Kronecker delta is
\[
  \delta_{ij}=\begin{cases}
  1,&i=j,\\
  0,&i\neq j.
  \end{cases}
\]
Thus the matrix \((\delta_{ij})\) is the identity matrix:
\[
  (\delta_{ij})=
  \begin{pmatrix}
  1&0&0\\
  0&1&0\\
  0&0&1
  \end{pmatrix}.
\]
\end{definition}

\section{Einstein summation convention}

\begin{definition}
Einstein summation convention says that when an index is repeated exactly twice in a term, summation over that index is understood:
\[
  x_iy_i=\sum_i x_iy_i.
\]
A repeated index is called a dummy index, and its name does not matter:
\[
  x_iy_i=x_jy_j=x_ky_k.
\]
\end{definition}

The basic rules are:
\begin{enumerate}[(i)]
  \item an index appearing once is a free index;
  \item an index appearing twice is summed over;
  \item an index appearing three or more times in one term is not allowed.
\end{enumerate}

\section{Rank}

\begin{definition}
The column rank of a matrix is the maximal number of linearly independent columns.  The row rank is the maximal number of linearly independent rows.  These two numbers are equal, and their common value is the rank of the matrix.
\end{definition}

Gaussian elimination is the standard way to compute rank: reduce the matrix to row echelon form and count the nonzero rows.


\section{Transpose and Hermitian conjugate}

\begin{definition}
If \(A\) is an \(m\times n\) matrix, its transpose \(A^T\) is the \(n\times m\) matrix defined by
\[
  (A^T)_{ij}=A_{ji}.
\]
\end{definition}

\begin{definition}
The Hermitian conjugate of \(A\) is
\[
  A^*=(A^T)^{\overline{\phantom{x}}}=\overline{A}^{\,T}.
\]
It satisfies
\[
  (AB)^*=B^*A^*.
\]
Indeed,
\[
  (AB)^*=\overline{(AB)}^T=(\overline A\,\overline B)^T=\overline B^T\overline A^T=B^*A^*.
\]
\end{definition}

\begin{definition}
A matrix is symmetric if
\[
  A^T=A.
\]
A matrix is Hermitian if
\[
  A^*=A.
\]
A matrix is skew-Hermitian if
\[
  A^*=-A.
\]
In a skew-Hermitian matrix, the diagonal entries are purely imaginary.
\end{definition}


\section{Trace}

\begin{definition}
The trace of an \(n\times n\) matrix \(A\) is the sum of its diagonal entries:
\[
  \operatorname{tr}(A)=\sum_i A_{ii}.
\]
In Einstein notation this is simply
\[
  \operatorname{tr}(A)=A_{ii}.
\]
\end{definition}


\section{A more structural view}

The common theme is index notation.  The Kronecker delta represents the identity, repeated indices encode summation, and matrix operations such as transpose, adjoint, and trace become compact when written with indices.

\section{Transpose and dual maps}

If $A:V\to W$ is a linear map, the dual map is
\[
  A^*:W^*\to V^*,\qquad A^*(\varphi)=\varphi\circ A.
\]
In bases, this dual map is represented by the transpose matrix.  This explains why transposition reverses order:
\[
  (AB)^T=B^TA^T.
\]
Composition of maps reverses when pulled back to functionals.

The elementary transpose is therefore not just flipping a matrix; it is the coordinate form of passing from vectors to covectors.

\section{Hermitian adjoints in Hilbert spaces}

In an inner product space, the adjoint $A^*$ is defined by
\[
  \langle Ax,y\rangle=\langle x,A^*y\rangle.
\]
Over $\mathbb C$, its matrix is the conjugate transpose.  A Hermitian matrix satisfies $A=A^*$ and has real eigenvalues.  A unitary matrix satisfies $A^*A=I$ and preserves the inner product.

The elementary real symmetric matrix is thus the real finite-dimensional case of self-adjoint operators.  This becomes central in quantum mechanics, Fourier analysis, and spectral theory.

\section{Trace as invariant}

The trace is the sum of diagonal entries:
\[
  \operatorname{tr}A=\sum_i a_{ii}.
\]
It is invariant under similarity because
\[
  \operatorname{tr}(P^{-1}AP)=\operatorname{tr}(APP^{-1})=\operatorname{tr}A.
\]
The cyclic rule
\[
  \operatorname{tr}(AB)=\operatorname{tr}(BA)
\]
is the key identity.

In category-theoretic language, trace abstracts the idea of closing an input-output wire.  In finite-dimensional vector spaces, that abstract trace reduces to the familiar matrix trace.  This is why traces appear in representation theory, statistical mechanics, and index theory.

\section{Schatten norms}

Rank counts nonzero singular values, while trace and determinant combine eigenvalues.  A unified family of matrix norms uses singular values:
\[
  \|A\|_{S_p}=\left(\sum_i \sigma_i(A)^p\right)^{1/p}.
\]
For $p=2$ this is the Frobenius norm; for $p=1$ it is the nuclear norm.  These norms are finite-dimensional models of operator ideals.

Thus rank, trace, Hermitian adjoint, and transpose are all pieces of a larger operator-theoretic vocabulary.

% Second-pass deepening for waves 2-4: wave 2
\section{Rank-nullity as a statement about information}

For a linear map $A:V\to W$, the kernel is the information lost by the map and the image is the information that remains visible.  Rank-nullity says
\[
  \dim V=\dim\ker A+\dim\operatorname{im}A.
\]
This is not just a dimension formula; it is an accounting law for information under a linear transformation.

The transpose gives the dual accounting.  The row space of a matrix consists of linear functionals on the input.  The null space consists of vectors annihilated by all those functionals.  Thus
\[
  \operatorname{Null}(A)=\operatorname{Row}(A)^\perp.
\]
Similarly,
\[
  \operatorname{Null}(A^T)=\operatorname{Col}(A)^\perp.
\]
These four subspaces are the geometric meaning of row reduction.

\section{Trace and representation characters}

If a group $G$ acts linearly on a vector space $V$, each group element gives a matrix $\rho(g)$.  The character of this representation is
\[
  \chi_\rho(g)=\operatorname{tr}(\rho(g)).
\]
Trace becomes a class function: conjugate group elements have the same trace because trace is invariant under similarity.  Thus the elementary identity
\[
  \operatorname{tr}(P^{-1}AP)=\operatorname{tr}(A)
\]
is the reason characters can classify representations of finite groups.

This gives a research-level continuation of a basic matrix statistic.  Trace is not only the sum of diagonal entries; it is the observable that remains after changing basis.

\section{Adjoints and energy}

For a Hermitian matrix $A$, the quadratic form
\[
  \langle Ax,x\rangle
\]
is real.  In quantum mechanics, Hermitian operators represent observables; in PDE, self-adjoint differential operators have real spectra; in numerical linear algebra, symmetric matrices admit stable orthogonal diagonalization.  The elementary condition $A^*=A$ therefore encodes a deep compatibility between algebra and geometry: the operator can be studied by orthogonal eigenmodes.
