\notechapter{N-20230310}{Derivatives, Tangents, Extremal Problems, Young--Holder--Minkowski, and Jensen}

\section{The derivative from secants to tangents}

The note begins with the geometric idea of derivative.  Given two nearby points on the curve,
\[
  (x_0,f(x_0)),\qquad (x_0+\Delta x,f(x_0+\Delta x)),
\]
the secant slope is
\[
  \frac{\Delta y}{\Delta x}
  =\frac{f(x_0+\Delta x)-f(x_0)}{\Delta x}.
\]
If, as $\Delta x\to0$, this quotient tends to a finite limit, then the curve has a definite tangent line at $x_0$.  The derivative is
\[
  f'(x_0)=\lim_{\Delta x\to0}
  \frac{f(x_0+\Delta x)-f(x_0)}{\Delta x}.
\]
Equivalently,
\[
  f'(x_0)=\lim_{x\to x_0}\frac{f(x)-f(x_0)}{x-x_0}.
\]

The tangent line at $x_0$ is therefore
\[
  y-f(x_0)=f'(x_0)(x-x_0).
\]
The source page stresses the visual transition: the secant line approaches a unique limiting line.  If that limiting line is not vertical and the slope is finite, we call the slope the derivative.

\begin{center}
\begin{tikzpicture}[scale=.78,>=Stealth,every node/.style={fill=white,inner sep=1pt}]
  \draw[->] (-1,0)--(7,0) node[right] {$x$};
  \draw[->] (0,-1)--(0,6) node[above] {$y$};
  \draw[domain=-.4:4.2,smooth,variable=\x] plot ({\x},{0.4*(\x-0.5)^2+1});
  \coordinate (A) at (1,1.1);
  \coordinate (B) at (3,3.5);
  \fill (A) circle (2pt);
  \node[below left=2pt] at (A) {$(x_0,f(x_0))$};
  \fill (B) circle (2pt);
  \node[above right=2pt] at (B) {$(x_0+\Delta x,f(x_0+\Delta x))$};
  \draw[dashed] (1,0)--(A);
  \draw[dashed] (3,0)--(B);
  \draw[dashed] (A)--(1,3.5)--(B);
  \draw[thick] (-.2,-.34)--(4.2,4.94);
  \node[right=2pt] at (4.2,4.94) {secant};
  \draw[thick] (-.1,.66)--(3,3.7);
  \node[above left=2pt] at (2.45,3.15) {tangent limit};
  \draw[<->] (1,-.35)--(3,-.35) node[midway,below] {$\Delta x$};
  \draw[<->] (3.25,1.1)--(3.25,3.5) node[midway,right] {$\Delta y$};
\end{tikzpicture}
\end{center}

\section{Derivative notation and local definition}

The definition requires $f$ to be defined on a punctured neighborhood of $x_0$, and usually on a small open interval around it.  The derivative at $x_0$ may be denoted by any of
\[
  f'(x_0),\qquad
  \left.\frac{df}{dx}\right|_{x=x_0},\qquad
  y'(x_0),\qquad
  \left.\frac{dy}{dx}\right|_{x=x_0}.
\]
If $f$ is differentiable at every point of an open interval $(a,b)$, then $x\mapsto f'(x)$ is the derivative function on that interval.

The note emphasizes a small logical point: before asking whether $f$ is differentiable at a point, one must first know that the function is defined near that point.  Differentiability is a local property.

\section{Rules of differentiation}

If $f$ and $g$ are differentiable at $x_0$, then sums, scalar multiples, products, and quotients obey:
\[
  (af+bg)'=af'+bg',\qquad
  (fg)'=f'g+fg',\qquad
  \left(\frac fg\right)'=\frac{f'g-fg'}{g^2},\quad g(x_0)\ne0.
\]
The chain rule says
\[
  (f\circ g)'(x_0)=f'(g(x_0))g'(x_0).
\]
The inverse-function rule says that if $f$ and $f^{-1}$ are inverse functions, $f'(x_0)\ne0$, and $y_0=f(x_0)$, then
\[
  (f^{-1})'(y_0)=\frac1{f'(x_0)}.
\]

These rules are not simply a table to memorize.  Each rule is a disguised statement about first-order linear approximation.  For example,
\[
  f(x_0+h)=f(x_0)+f'(x_0)h+o(h),\qquad
  g(x_0+h)=g(x_0)+g'(x_0)h+o(h),
\]
and multiplying these expansions gives the product rule.

\section{Elementary derivative formulas}

The table in the note includes the standard formulas:
\[
  C'=0,\qquad (x^a)'=ax^{a-1},\qquad (a^x)'=a^x\ln a,\qquad
  (\log_a x)'=\frac1{x\ln a}.
\]
For trigonometric functions,
\[
  (\sin x)'=\cos x,\qquad
  (\cos x)'=-\sin x,\qquad
  (\tan x)'=\sec^2x,\qquad
  (\cot x)'=-\csc^2x,
\]
\[
  (\sec x)'=\sec x\tan x,\qquad
  (\csc x)'=-\csc x\cot x.
\]
For inverse trigonometric functions,
\[
  (\arcsin x)'=\frac1{\sqrt{1-x^2}},\qquad
  (\arccos x)'=-\frac1{\sqrt{1-x^2}},\qquad
  (\arctan x)'=\frac1{1+x^2}.
\]

\section{Extrema, Fermat's lemma, and critical points}

A point $x_0$ is a local maximum if $f(x)\le f(x_0)$ near $x_0$, and a local minimum if $f(x)\ge f(x_0)$ near $x_0$.  If $f'(x_0)=0$, then $x_0$ is called a stationary or critical point.

Fermat's lemma says: if $f$ is differentiable at $x_0$ and has a local extremum there, then
\[
  f'(x_0)=0.
\]
The note writes a warning beside this theorem: the converse is false.  For example,
\[
  f(x)=x^3
\]
has $f'(0)=0$, but $0$ is not an extremum.  Fermat's lemma only provides candidates for extrema.  One must then check sign changes, second derivatives, endpoints, or non-differentiable points.

If $f''(x_0)>0$ and $f'(x_0)=0$, then $x_0$ is a local minimum.  If $f''(x_0)<0$, it is a local maximum.  The reason is that $f''$ describes whether $f'$ is increasing or decreasing near the critical point.

\section{A derivative proof of a basic inequality}

The note proves, for $x>0$,
\[
  x^\alpha-\alpha x+\alpha-1\le0,\qquad 0<\alpha<1,
\]
and
\[
  x^\alpha-\alpha x+\alpha-1\ge0,\qquad \alpha<0\text{ or }\alpha>1.
\]
Let
\[
  F(x)=x^\alpha-\alpha x+\alpha-1.
\]
Then
\[
  F'(x)=\alpha(x^{\alpha-1}-1),\qquad F(1)=0.
\]
For $0<\alpha<1$, $F$ increases on $(0,1)$ and decreases on $(1,\infty)$, so $x=1$ is a maximum and $F(x)\le0$.  For $\alpha<0$ or $\alpha>1$, the monotonicity is reversed, and $x=1$ is a minimum, so $F(x)\ge0$.

This small inequality is the engine behind Young's inequality.

\section{Young's inequality}

Let $a,b>0$ and let $p,q$ satisfy
\[
  \frac1p+\frac1q=1.
\]
For $p>1$, Young's inequality says
\[
  a^{1/p}b^{1/q}\le \frac ap+\frac bq,
\]
with equality iff $a=b$.  The proof substitutes
\[
  x=\frac ab,\qquad \alpha=\frac1p
\]
into the previous derivative inequality, then multiplies by $b$.

The source also records the reversed version for $p<1$, where the inequality sign changes.  The important conceptual point is that an inequality between weighted geometric and arithmetic means is reduced to the shape of one single-variable function.

\section{Holder's inequality}

For $x_i,y_i\ge0$ and $1/p+1/q=1$, Holder's inequality is
\[
  \sum_{i=1}^n x_i y_i
  \le
  \left(\sum_{i=1}^n x_i^p\right)^{1/p}
  \left(\sum_{i=1}^n y_i^q\right)^{1/q},\qquad p>1.
\]
The proof in the source normalizes
\[
  a_i=\frac{x_i^p}{\sum_j x_j^p},\qquad
  b_i=\frac{y_i^q}{\sum_j y_j^q}
\]
and applies Young's inequality termwise.  After adding the inequalities, the normalized sum becomes $1$, giving Holder.

A handwritten note observes that the integral form follows the same idea:
\[
  \int_a^b f(x)g(x)\,dx
  \le
  \left(\int_a^b |f(x)|^p\,dx\right)^{1/p}
  \left(\int_a^b |g(x)|^q\,dx\right)^{1/q}.
\]
The comment says this feels almost trivial once the finite-sum proof is understood: one replaces sums by integrals and uses the same normalization logic.

For $p=q=2$, Holder becomes the Cauchy--Bunyakovsky--Schwarz inequality.

\section{Minkowski's inequality}

For $p>1$, Minkowski's inequality says
\[
  \left(\sum_{i=1}^n (x_i+y_i)^p\right)^{1/p}
  \le
  \left(\sum_{i=1}^n x_i^p\right)^{1/p}
  +
  \left(\sum_{i=1}^n y_i^p\right)^{1/p}.
\]
The proof applies Holder to the two sums
\[
  \sum_i x_i(x_i+y_i)^{p-1},\qquad
  \sum_i y_i(x_i+y_i)^{p-1}.
\]
Because $q=p/(p-1)$, we have
\[
  ((x_i+y_i)^{p-1})^q=(x_i+y_i)^p.
\]
After factoring the common term and dividing by it, one obtains Minkowski.

The handwritten margin notes that for $p=2$ this is exactly the triangle inequality for Euclidean length:
\[
  \sqrt{(x_1+y_1)^2+\cdots+(x_n+y_n)^2}
  \le
  \sqrt{x_1^2+\cdots+x_n^2}+
  \sqrt{y_1^2+\cdots+y_n^2}.
\]

\section{Convex functions and Jensen's inequality}

The note defines a convex function by the inequality
\[
  f(\alpha_1x_1+\alpha_2x_2)
  \le
  \alpha_1f(x_1)+\alpha_2f(x_2),\qquad
  \alpha_1+\alpha_2=1,\quad \alpha_i\ge0.
\]
The handwritten annotation explains the geometry: the point on the chord between $(x_1,f(x_1))$ and $(x_2,f(x_2))$ lies above the graph.

Equivalently, for $x_1<x<x_2$,
\[
  \frac{f(x)-f(x_1)}{x-x_1}
  \le
  \frac{f(x_2)-f(x)}{x_2-x}.
\]
This slope inequality implies that if $f$ is differentiable, then $f'$ is increasing; if $f$ is twice differentiable, then convexity corresponds to $f''\ge0$.

Jensen's inequality generalizes the two-point definition:
\[
  f\left(\sum_{i=1}^n \alpha_i x_i\right)
  \le
  \sum_{i=1}^n \alpha_i f(x_i),\qquad
  \alpha_i\ge0,\quad \sum_i\alpha_i=1.
\]
The source proves this by induction.  Let $\beta=\alpha_2+\cdots+\alpha_n=1-\alpha_1$ and normalize the remaining weights by $\alpha_i/\beta$.  Then use the two-point convexity inequality followed by the induction hypothesis.

\section{Typical exercise themes}

The last pages list exercises.  They are not random; they are organized around the derivative as a tool for inequality and extrema.

Examples include:
\begin{enumerate}[(i)]
  \item determining when a line and cubic have three distinct intersections;
  \item finding monotonicity intervals of $f(x)=\sqrt x-\ln(x+a)$;
  \item maximizing $\sin^2x\cos x$ on $(0,\pi/2)$;
  \item comparing numbers such as $0.1e^{0.1}$, $1/9$, and $-\ln0.9$ using auxiliary functions;
  \item proving $e^x\ge1+x$ and $\ln x\le x-1$;
  \item using Jensen or logarithms to prove entropy-like inequalities such as
  \[
    \sum_{i=1}^{2^n}p_i\log_2p_i\ge -n,\qquad \sum_i p_i=1;
  \]
  \item proving product inequalities of the form
  \[
    \prod_{i=1}^n a_i^{b_i}\le1
  \]
  under weighted-sum hypotheses;
  \item minimizing expressions such as
  \[
    \frac9{\sin^2x}+\frac1{\cos x\cos y\sin^2y}.
  \]
\end{enumerate}

The common method is to build an auxiliary function, compute derivatives, locate monotonicity or convexity, and then read off the desired inequality.

\section{The reusable pattern}

This note shows how the derivative becomes more than a tangent slope.  It gives a language for extrema, monotonicity, convexity, and inequalities.  Young, Holder, Minkowski, and Jensen all look like different inequalities, but the note reveals a chain:
\[
\begin{gathered}
  \text{single-variable derivative inequality}\to \text{Young}\to \text{Holder}\\
  \to \text{Minkowski},\qquad \text{convexity}\to \text{Jensen}.
\end{gathered}
\]
Thus calculus becomes a proof technology for analysis and inequalities.


\paragraph{Expanded synthesis.}
The common theme of this chapter is that an elementary limiting or computational rule becomes powerful only after one specifies the space in which the limiting process takes place. The earlier sections (`The derivative from secants to tangents`, `Derivative notation and local definition`, `Rules of differentiation`, `Elementary derivative formulas`, `Extrema, Fermat's lemma, and critical points`) may look like separate techniques, but they all ask the same question: which operations are stable under approximation? A derivative is stable first-order approximation,
\[
  f(a+h)=f(a)+L(h)+o(|h|),
\]
while an integral is stable accumulation with respect to a measure, and a convergent series is a limit in a chosen norm or topology.

The advanced bridge is functional-analytic. Uniform convergence is convergence in a sup norm; $L^p$ convergence is convergence in an integral norm; differentiability is the existence of a continuous linear approximation; many differential equations are operator equations $L[u]=f$. Theorems such as dominated convergence, the inverse function theorem, the projection theorem in Hilbert spaces, and semigroup existence results are refined answers to the same elementary concern: when may we pass from finite approximations to a genuine limiting object?

To use this viewpoint when rereading the original examples, identify the hidden stability condition. A Taylor expansion asks for control of the remainder. A change of variables asks for differentiability and Jacobian control. Termwise differentiation of a series asks for uniform convergence of derivative series. A differential-equation formula asks whether the operator is invertible under the imposed initial or boundary data. The elementary computation is the visible part; the advanced theory names the hypotheses that make it legitimate.
