\notechapter{N-20220713}{Completeness, Supremum, Dedekind Cuts, and Metric Spaces}

\section{Large and small positive real numbers}

The page begins with a simple but important fact: for every positive real \(A\), there is a positive real \(M>A\); for every positive real \(a\), there is a positive real \(\varepsilon<a\).  For instance,
\[
  M=A+1,\qquad \varepsilon=\frac a2.
\]
This is the first appearance of the \(\varepsilon\)-language of analysis.

\section{The geometric view of the real line}

We think of \(\mathbb R\) as a line, where order is represented by left and right.  This picture is helpful, but the note emphasizes that all conclusions must ultimately come from axioms and rigorous reasoning, not from the drawing itself.

A set \(X\subseteq\mathbb R\) is bounded above if there exists \(A\in\mathbb R\) such that
\[
  x\le A\quad\text{for all }x\in X.
\]
Such an \(A\) is an upper bound.  Similarly, \(X\) is bounded below if there exists \(a\) such that
\[
  a\le x\quad\text{for all }x\in X.
\]

\section{Supremum property}

The least upper bound property says: if \(X\subseteq\mathbb R\) is nonempty and bounded above, then the set of upper bounds has a least element.  This least upper bound is denoted
\[
  \sup X.
\]
Equivalently, \(M=\sup X\) iff:
\begin{enumerate}[(i)]
  \item \(x\le M\) for all \(x\in X\);
  \item for every \(\varepsilon>0\), there exists \(x\in X\) such that
  \[
    x>M-\varepsilon.
  \]
\end{enumerate}
The note sketches a proof of the supremum property from the nested interval principle by bisecting intervals and selecting the half that still contains upper bounds or points of the set.

The dual notion is the infimum:
\[
  \inf X.
\]
One always has
\[
  \inf X\le \sup X
\]
for a nonempty bounded set.

\section{Nested intervals imply completeness}

Let \(X\subseteq\mathbb R\) be nonempty and bounded above.  One can construct nested intervals
\[
  I_n=[a_n,b_n]
\]
so that \(a_n\) is not an upper bound of \(X\), while \(b_n\) is an upper bound, and
\[
  b_n-a_n\le 2^{-n}.
\]
The nested interval axiom gives a point
\[
  M\in\bigcap_{n=1}^{\infty}I_n.
\]
The shrinking length forces \(M\) to be the least upper bound of \(X\).  If \(M\) were not an upper bound, some \(x\in X\) would lie above it; for large \(n\), the right endpoint \(b_n\) would be too close to \(M\), contradiction.  If \(M\) were not least, a smaller upper bound would contradict the construction of the \(a_n\)'s.

\section{Dedekind cuts}

A Dedekind cut is a subset \(X\subset\mathbb Q\) such that:
\begin{enumerate}[(i)]
  \item \(X\neq\varnothing\) and \(\mathbb Q\setminus X\neq\varnothing\);
  \item if \(x\in X\) and \(y<x\), then \(y\in X\);
  \item \(X\) has no greatest element.
\end{enumerate}
A rational number \(p/q\) defines the cut
\[
  X_{p/q}=\{x\in\mathbb Q:x<p/q\}.
\]
An irrational such as \(\sqrt2\) is represented by
\[
  X_{\sqrt2}=\{x\in\mathbb Q_{>0}:x^2<2\}\cup\mathbb Q_{\le0}.
\]
Thus real numbers can be constructed from rational cuts.

\section{Metric spaces}

A metric space is a set \(X\) equipped with a function
\[
  d:X\times X\to\mathbb R_{\ge0}
\]
satisfying:
\begin{enumerate}[(i)]
  \item \(d(x,y)=0\) iff \(x=y\);
  \item \(d(x,y)=d(y,x)\);
  \item \(d(x,z)\le d(x,y)+d(y,z)\).
\end{enumerate}

The usual metric on \(\mathbb R\) is
\[
  d(x,y)=|x-y|.
\]
On \(\mathbb R^q\), common metrics are:
\[
  d_2(x,y)=\left(\sum_{n=1}^q |x_n-y_n|^2\right)^{1/2},
\]
\[
  d_1(x,y)=\sum_{n=1}^q |x_n-y_n|,
\]
and
\[
  d_\infty(x,y)=\max_{1\le n\le q}|x_n-y_n|.
\]
The note remarks that \(d_2\) requires Cauchy--Schwarz for the triangle inequality, \(d_1\) is the Manhattan metric, and \(d_\infty\) is often called the Chebyshev metric.

A final example is the discrete metric:
\[
  d(x,y)=
  \begin{cases}
  0,&x=y,\\
  1,&x\neq y.
  \end{cases}
\]
In the discrete metric, open balls are especially simple:
\[
  B_r(x)=
  \begin{cases}
  \{x\},&0<r\le1,\\
  X,&r>1.
  \end{cases}
\]
The note points out that boundary points are not included in an open ball, explaining why the case \(r=1\) gives \(\{x\}\), not all of \(X\).

\section{Subspaces and density}

If \(Y\subseteq X\) and \(X\) has metric \(d\), then \(Y\) inherits the subspace metric
\[
  d_Y(y_1,y_2)=d(y_1,y_2).
\]
A subset \(Y\subseteq X\) is dense if for every \(x\in X\) and every \(\varepsilon>0\), there exists \(y\in Y\) such that
\[
  d(x,y)<\varepsilon.
\]
This formalizes the idea that \(Y\) comes arbitrarily close to every point of \(X\).


\section{Supremum as least upper bound}

The supremum of a set \(A\subseteq\mathbb R\) is not merely a maximum.  It is the least number \(s\) such that every element of \(A\) is at most \(s\).  A maximum must belong to \(A\); a supremum need not.

For example,
\[
  \sup(0,1)=1,
\]
but \((0,1)\) has no maximum.  This distinction is exactly what makes completeness subtle.

\section{Dedekind cuts and completeness}

A Dedekind cut separates rationals into a lower part and an upper part with no gap.  Irrational numbers appear as cuts not produced by a rational boundary.  Completeness says that every such cut corresponds to a real number.

This is the philosophical point of constructing \(\mathbb R\): the real line is obtained by filling all rational gaps.

\section{Expanded synthesis and advanced context}

The note moves from the order-completeness of \(\mathbb R\) to the metric-space language of distance and density.  Supremum, Dedekind cuts, nested intervals, and metrics are different ways of making precise the intuition that the real line has no holes.


\paragraph{Expanded synthesis.}
The higher-level content of this chapter is the passage from pointwise description to invariant structure. The earlier sections (`Large and small positive real numbers`, `The geometric view of the real line`, `Supremum property`, `Nested intervals imply completeness`, `Dedekind cuts`) introduce objects that may be drawn, listed, glued, or computed, but the advanced question is what remains unchanged under the allowed maps. In topology the allowed maps are continuous maps; in category theory they are morphisms preserving the relevant structure; in homology they induce maps between algebraic invariants.

The guiding pattern is
\[
  \text{space or structure}\quad \xrightarrow{\text{allowed maps}}\quad \text{invariant data}.
\]
Quotient spaces formalize gluing; products formalize simultaneous coordinates; compactness formalizes the impossibility of escaping by a limiting process; connectedness formalizes the impossibility of separation; homology converts holes into groups. When a note discusses closure, quotienting, paths, products, or functors, it is already working with this invariant language.

A quotient exercise is a gluing construction; a compactness exercise is an infinite pigeonhole argument; a homotopy exercise is a controlled deformation; a functorial example says that a construction is compatible with maps. The advanced theory does not add decoration. It explains why the same few operations, such as taking products, quotients, images, kernels, and induced maps, occur throughout topology, algebra, and geometry.


\section{Pedagogical repair: the many faces of completeness}

Supremum, nested intervals, Dedekind cuts, and Cauchy completeness look different at first, but they all express the same intuition: the real line has no holes.  The note is valuable precisely because it touches several versions of this intuition.

The least-upper-bound property is order-theoretic; nested intervals are geometric; Dedekind cuts are constructive; metric completeness is analytic.  Later analysis freely moves among these viewpoints.  A sequence proof may secretly use a supremum; a construction of \(\sqrt2\) may use a cut; a fixed-point theorem may use metric completeness.
