
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book is a reorganized edition of my informal mathematical notes, written across a long stretch of time from middle school to the present. In that sense, \emph{INM} is not only a mathematical document, but also a personal record. It preserves the way a subject first appeared to me, the way a computation was once hard, the way a theorem gradually stopped being a name and became a tool, and the way one topic kept opening a door into another.

The original notes were not planned as a textbook. They were accumulated from exercises, self-study, lectures, fragments of reading, exam preparation, conversations, and attempts to explain ideas to myself. Some entries begin with very elementary problems: set identities, inequalities, Euclidean geometry, congruences, derivatives, determinants, or multivariable integrals. Other entries move toward more advanced objects: quotient spaces, exact sequences, cyclotomic fields, exterior algebra, Hilbert spaces, differential forms, de Rham cohomology, Fourier expansions, operator methods, topos-flavored foundations, mathematical physics, and the mathematics around machine learning. The purpose of this edition is to let these layers coexist rather than to hide one behind the other.

A recurring intention in the reorganization is to connect elementary techniques with the higher mathematics they quietly anticipate. A Venn diagram is also Boolean algebra and measurable-event logic. A gcd computation is also ideal arithmetic in a principal ideal domain. A determinant is also the action on the top exterior power. A line integral is also the pullback of a differential form. Fourier coefficients are also Hilbert-space coordinates. These connections are not meant to make the elementary material look more sophisticated than it is; they are meant to show why the elementary material is worth keeping. The first calculation is often the most honest place where a later abstraction can be seen.

The present volume arrangement follows the main themes of the original notes. The book is divided into Foundations, Arithmetic and Algebra, Combinatorics, Linear Algebra, Calculus and Inequalities, Multivariable Analysis, Fourier and ODEs, Geometry, Topology, and Research Vignettes. The last part is deliberately kept for material that is exploratory, cross-disciplinary, or difficult to classify cleanly: small mathematical-physics notes, AI-related mathematical explanations, and research-flavored digressions. I prefer to keep such pieces visible, because they record moments when mathematics was not a syllabus but a landscape.

There is also a memorial reason for making this collection. A notebook from one's earlier years is full of unevenness: some pages are naive, some are overambitious, some are clumsy, and some still contain the excitement of first contact. To reorganize these notes is not to erase that unevenness, but to give it form. The book marks a path from contest-style manipulation and classroom calculus toward algebra, analysis, topology, geometry, and the modern languages that now feel unavoidable. It is a record of mathematical growth, but also of the persistence of certain questions: what is the right structure, what is invariant, what is being measured, and why should one care?

This edition is therefore best read as a working companion rather than a polished authority. Definitions and theorems are included where they clarify the original notes, but the spirit remains informal. I have tried to preserve computations, examples, and small motivations, while adding enough surrounding structure to make the notes reusable. Some chapters are closer to textbook exposition; others are closer to annotated memory. That mixture is intentional. It is the shape of the archive.

\begin{flushright}
Runnel Zhang\\
\today
\end{flushright}
